Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-14 WSSSPE1_report/WSSSPE1_summary_report.tex
Add non-breaking spaces and other minor changes.

\subsection{Training Scientists to Develop and Support Software}
Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work. 
One question raised was whether a software developer can be effective if they are not part
of the appropriate domain community, with responses that this depends on the 
specific problem and situation, and that "T-shaped" people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
The discussion focused on  whether we could teach software developers and domain scientists to 
collaborate together more effectively rather than trying to teach software developers
about domain science and domain scientists about software development practices. 
The end goal of this would be to have a single community with a spectrum of expertise across
domain science and software development rather than two separate communities.

@@

There was debate that while some of the basics of computing may be 
similar tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
 There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

The workshop submissions grouped under this section provide frameworks
for thinking about, researching and understanding which elements of sustainability
are important and how they relate. Although there is substantial
overlap between the frameworks, they have different emphases and
extents.  Each paper in this group included a definition of sustainability,
with many overlaps.  In essence the papers related the question to
change over time, an idea well captured in the definition used by the
UK's Scientific Software Sustainability Institute: ``software you use
today will be available---and continue to be improved and
supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
idea was also expressed as software that continues to serve its
users~\cite{Pierce_WSSSPE}.  For a more thorough discussion of definitions
of software sustainability see \S\ref{sec:defining-sustainability}.

@@

Frameworks also concerned themselves with the context in which software
projects exist, moving in abstraction from the software artifact
itself to the organization of its production and
the shape of the space in which an artifact or project exists.
These frameworks take the form of contingency theories in that
they outline a different set of challenges and argue that different
project organizations (and presumably artifact attributes) are
necessary to persist longterm in spaces with particular
characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
of this idea of space in terms of three axes: temporal (long or short
term needs), spatial (local or global use) and purpose (specific to
general).  They propose that different project organizations will be
needed in different locations and argue that we should concentrate
research to understand those connections. Weber et
al.~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
ecosystems as ``niches'' which sustain particular pieces of
software. They define ``a software niche as the set of technical
requirements, organizational conditions and cultural mores that
support its maintenance and use over time.'' They call for better
understanding and modeling of niches (as well as further exploration
of the usefulness of ecosystem metaphors).

@@

The discussion of development-focused communities centered around describing
methods of interaction between individuals and the scientific software.  The
first type of interaction was the development of a specific piece of software,
the second was a particular domain or discipline, and the final primary
type of interaction was around the development of applications built on a
particular piece of software that was perhaps developed by another group.  As
an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
both the for-profit company \toolname{Kitware} and the users and contributors to their
software packages such as \toolname{VTK}.  This structure, of the centralized development
of core infrastructure around which communities of individuals applying that
infrastructure and developing applications utilizing it, was similarly
reflected in \cite{Terrel_WSSSPE}, where the core scientific python ecosystem
is supported by a non-profit entity that fosters community investment in that
ecosystem.  In many ways, these two organizations (\toolname{Kitware} and \toolname{NumFOCUS})
attempt to cross domain boundaries and provide support for both the
infrastructure and application sides of community building.

@@

Both \cite{howison_scientific_2011} and \cite{howison_incentives_2013} 
study how scientific software collectives self-organize and address issues of
incentive, long-term support, and development of infrastructure as well as new
features.  As noted elsewhere in this summary,
\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
communities in astrophysics.

From the perspective of developing prescriptions for successful scientific
software development, both \cite{citeulike:11831265} and
\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
for developing open source scientific software, focusing on both the choices
made during development and the sustainability of practices in the long term.  
\cite{1749-4699-6-1-015010} describes the development and long-term growth of
the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
applications and domains has shaped its development and community trajectory.

From more traditional open source development, resources were shared that
developed communities explicitly, such as 
\cite{citeulike:7888211} and \cite{Trapani:2011}, focusing on
large-scale projects such as the Ubuntu linux distribution and smaller-scale
volunteer-developed projects like such as ThinkUp, respectively.
The process of open source development, while less explicitly focused on
community building, outlined in \cite{citeulike:478633} was seen as a valuable
resource, particularly when combined with the management and personal
interaction techniques outlined in \cite{opac-b1134063}.  Growing diversity in
communities was directly addressed in \cite{Allsopp:2012}, where experiences
growing the diversity of technical conferences in open source were described.

>>>>

\subsection{Training Scientists to Develop and Support Software}
Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work. 
One question raised was whether a software developer can be effective if they are not part
of the appropriate domain community, with responses that this depends on the 
specific problem and situation, and that ``T-shaped'' people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
The discussion focused on  whether we could teach software developers and domain scientists to 
collaborate together more effectively rather than trying to teach software developers
about domain science and domain scientists about software development practices. 
The end goal of this would be to have a single community with a spectrum of expertise across
domain science and software development rather than two separate communities.

@@

There was debate that while some of the basics of computing may be 
similar, tools commonly used for computing have changed quite
significantly. Reference was made to Perl, which was commonly used, giving way
to whole new languages, such as Python, for gluing things together and how this
induces many students into entirely rewriting the scaffolding, leaving the old
to rot and the experiments to become non-reproducible as the tools change.
 There was
discussion of this tendency along with the enormous differences in the speed and
ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
the immediate sharing of the latest development online, using revision control
systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.

@@

The workshop submissions grouped under this section provide frameworks
for thinking about, researching and understanding which elements of sustainability
are important and how they relate. Although there is substantial
overlap between the frameworks, they have different emphases and
extents.  Each paper in this group included a definition of sustainability,
with many overlaps.  In essence the papers related the question to
change over time, an idea well captured in the definition used by the
UK's Scientific Software Sustainability Institute: ``software you use
today will be available---and continue to be improved and
supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
idea was also expressed as software that continues to serve its
users~\cite{Pierce_WSSSPE}.  For a more thorough discussion of definitions
of software sustainability see~\S\ref{sec:defining-sustainability}.

@@

Frameworks also concerned themselves with the context in which software
projects exist, moving in abstraction from the software artifact
itself to the organization of its production and
the shape of the space in which an artifact or project exists.
These frameworks take the form of contingency theories in that
they outline a different set of challenges and argue that different
project organizations (and presumably artifact attributes) are
necessary to persist long term in spaces with particular
characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
of this idea of space in terms of three axes: temporal (long or short
term needs), spatial (local or global use) and purpose (specific to
general).  They propose that different project organizations will be
needed in different locations and argue that we should concentrate
research to understand those connections. Weber et
al.~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
ecosystems as ``niches'' which sustain particular pieces of
software. They define ``a software niche as the set of technical
requirements, organizational conditions and cultural mores that
support its maintenance and use over time.'' They call for better
understanding and modeling of niches (as well as further exploration
of the usefulness of ecosystem metaphors).

@@

The discussion of development-focused communities centered around describing
methods of interaction between individuals and the scientific software.  The
first type of interaction was the development of a specific piece of software,
the second was a particular domain or discipline, and the final primary
type of interaction was around the development of applications built on a
particular piece of software that was perhaps developed by another group.  As
an example, in~\cite{Hanwell_WSSSPE}, the community described is comprised of
both the for-profit company \toolname{Kitware} and the users and contributors to their
software packages such as \toolname{VTK}.  This structure, of the centralized development
of core infrastructure around which communities of individuals applying that
infrastructure and developing applications utilizing it, was similarly
reflected in~\cite{Terrel_WSSSPE}, where the core scientific python ecosystem
is supported by a non-profit entity that fosters community investment in that
ecosystem.  In many ways, these two organizations (\toolname{Kitware} and \toolname{NumFOCUS})
attempt to cross domain boundaries and provide support for both the
infrastructure and application sides of community building.

@@

Both~\cite{howison_scientific_2011} and~\cite{howison_incentives_2013} 
study how scientific software collectives self-organize and address issues of
incentive, long-term support, and development of infrastructure as well as new
features.  As noted elsewhere in this summary,
\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
communities in astrophysics.

From the perspective of developing prescriptions for successful scientific
software development, both~\cite{citeulike:11831265} and
\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
for developing open source scientific software, focusing on both the choices
made during development and the sustainability of practices in the long term.  
\cite{1749-4699-6-1-015010} describes the development and long-term growth of
the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
applications and domains has shaped its development and community trajectory.

From more traditional open source development, resources were shared that
developed communities explicitly, such as~\cite{citeulike:7888211} 
and~\cite{Trapani:2011}, focusing on
large-scale projects such as the Ubuntu linux distribution and smaller-scale
volunteer-developed projects like such as ThinkUp, respectively.
The process of open source development, while less explicitly focused on
community building, outlined in~\cite{citeulike:478633} was seen as a valuable
resource, particularly when combined with the management and personal
interaction techniques outlined in~\cite{opac-b1134063}.  Growing diversity in
communities was directly addressed in~\cite{Allsopp:2012}, where experiences
growing the diversity of technical conferences in open source were described.
